### 2018-05-06 4_impl.tex
add up to preproc.

\vspace{0.5em}
\begin{enumerate}[nolistsep]
  \item \textit{Robustness:}
    \begin{enumerate}[label*=\arabic*.]
      \item usage of immutable data structures for all DTOs;
      \item interrupting the work on errors;
      \item modular architecture: each module can be tested independently;
      \item usage of software design patterns if necessary;
    \end{enumerate}
  \item \textit{Transparency:}
    \begin{enumerate}[label*=\arabic*.]
      \item following the principles of simplicity and readability;
      \item clear and informative program output;
    \end{enumerate}
  \item \textit{Efficiency:}
    \begin{enumerate}[label*=\arabic*.]%[leftmargin=1.5em]
      \item keeping the trade-off between execution time and memory usage;
    \end{enumerate}
  \item \textit{Extensibility:}
    \begin{enumerate}[label*=\arabic*.]%[leftmargin=1em]
      \item clear modular architecture.
    \end{enumerate}
\end{enumerate}

As \porthos[1], the \porthos[2] uses the open-source SMT solver \texttt{Z3} from Microsoft Research~\cite{de2008z3}. However, unlikely its predecessor, the \porthos[2] has an additional abstraction level \zformula{} (see Section~\ref{ch:impl:model:zformula}) that allow to use any other SMT solver. %TODO: *should* allow?

@@

Each memory unit has an \texttt{XType} associated with it.
The X-type is a symbolic representation of the C primitive type%
\footnote{Here we should note the \porthos[2] can eventually evolve to be able to analyse programs written in an OOP language (for instance, in C++). In this case, the \texttt{XType} will have more complex structure than a simple enumeration, which it has when we need to emulate only primitive types of C language. See more detailed discussion on input language type system in Section~\ref{ch:impl:proc:x-pre-compiler:type}.} %
that is easily convertable to a SMT type (modelled as \texttt{ZType}).%TODO: add the class ZType

@@

  \1 \texttt{XComputationEvent} \\
    We distinguish two types of computation events:
      %\begin{itemize}[noitemsep]
      \2 \texttt{XUnaryComputationEvent} that encodes bit negation and no-operation,
      \2 \texttt{XBinaryComputationEvent} that encodes numeric operations (such as addition, multiplication, etc.), bit vector operations (such as bit-and, bit-xor, etc.), relative operations (such as greater-then comparison, equality comparison, etc.), and logical operations (such as conjunction and disjunction).
      %\end{itemize}
    Since a computation event performs computation over local-only memory and does not change value of any memory unit, the \textit{computation} abstraction (as the CPU time spent for the computation itself) can be safely removed from the model, and the computation event can be seen as a simultaneous operation.
    Therefore the computation event class implements both \texttt{XEvent} and \texttt{XMemoryUnit}.

  \1 \texttt{XControlFlowEvent} \\
    The control-flow event indicates a non-linear jump in the code. We distinguish two kinds of control-flow events: 
    %\begin{itemize}[noitemsep]
      \2 \texttt{XJumpEvent} that performs no computation and no data operation, it can be safely removed from the model as an optimisation,
      \2 \texttt{XMethodCallEvent} that models the method call with the \texttt{FASTCALL} calling convention (passing arguments in registers).
      %TODO: check the diagrams if the XMCE really inherits XLMU
      The method call also implements the \texttt{XLocalMemoryUnit} since it represents the computed value returned by the method call.
      If the method has been resolved, the actual arguments are bind to the formal parameters (treated as temporary local memory units), the called \texttt{XMethodCallEvent} is pushed onto the call stack, and the execution jumps to the method body.
      Each \texttt{return} statement creates the assignment of the method call event on the top of call stack.
      \\
      If the method has not been resolved, \textit{TODO: think about it. Will the formula be SAT? Or do we need to abort the analysis?}
    %\end{itemize}

@@

\begin{figure}[h]
\centering
\begin{multicols}{2}
\begin{lstlisting}[language=java]
interface Element {
  ...
  <T> T accept(Visitor<T> visitor);
}

class AnElement implements Element {
  ...
  @Override
  <T> T accept(Visitor<T> visitor) {
    return visitor.visit(this);
  }
}
[* \vfill\null \columnbreak *]
class Visitor<T> {
  ...

  T visit(AnElement e) {
    // some visiting logic
    ...
    // continue recursively
    e.getChild().accept(this);
  }

  T visit(AnOtherElement e) {
    ...
  }
}
\end{lstlisting}
\end{multicols}
\caption{Sketch of the visitor pattern}
\label{fig:visitor}
\end{figure}

@@

Both input-language and input-model parsers are implemented via ANTLR parser generator.
Details on the input program language grammar are discussed in Section~\ref{ch:impl:input}.
The input memory model language \cat is discussed in Section~\ref{ch:wmm:cat}.
The ANTLR grammar for the \cat language was extracted from the parser used by \tool{herd} tool%
\footnote{\tool{herd} project repository: \url{https://github.com/herd/herdtools7}} %
written in OCaml.
%TODO: upload the grammar to github
%The input memory model

\textbf{TODO: add reference to the grammar file in github repository OR include a short version as an appendix}

@@

todo: grammar of w-model to appendix

@@

\paragraph{The macro preprocessor}

The preprocessor traverses the \ytree{} and collects information necessary for compiling it into an \xgraph{}.

typedefs

aliases

more macros?

not supported yet

@@

@@

The label resolution is a process of linkage the referenced labels to declared labels.

In C, the labeled statements are declared via the colon-syntax `\texttt{<label> : <statement>}',
and the labels are referenced by the jump-statement `\texttt{goto <label>}'.
The label resolution algorithm traverses the \ytree{} and collects all declared labels into a map that points a label to the labeled statement. %TODO: code: remove temp label generation, it's stupid. put null.
This information is used during compilation (see Section~\ref{ch:impl:proc:x-compiler}).

\paragraph{Type analysis}% and function resolution.}
\label{ch:impl:proc:x-pre-compiler:type}

%TODO: remake this paragraph considering that we have already said sth about XType
The C language has a static (resolved at compile-time) manifest (all types are declared explicitly) type system.
Comparing to languages that use type inference, the type analysis of a C program constitutes a simple propagating the type information (obtained from variables declarations) to all expressions.
%In \porthos[2], the type is modelled by the class \texttt{XType}.%TODO: + fix class and package name!
Being carried at Y and X representation levels, the type is converted to a Z-type at the Z-formula encoding stage (see Section~\ref{ch:impl:proc:z-encoder}).

%TODO: describe the Type
%TODO: maybe rename back to XType
%TODO: but definitely get rid of bit32

@@

>>>>

\vspace{0.5em}
\begin{enumerate}[nolistsep]
  \item \textit{Robustness:}
    \begin{enumerate}[label*=\arabic*.]
      \item prohibit under-approximating abstractions;
      \item preservation the completeness of analysis;
      \item modular architecture: each module can be tested independently;
      \item usage of software design patterns where necessary;
      \item usage of immutable data structures for all DTOs;
    \end{enumerate}
  \item \textit{Transparency:}
    \begin{enumerate}[label*=\arabic*.]
      \item following the principles of simplicity and readability;
      \item clear and informative program output;
    \end{enumerate}
  \item \textit{Efficiency:}
    \begin{enumerate}[label*=\arabic*.]%[leftmargin=1.5em]
      \item keeping the trade-off between execution time and memory usage;
    \end{enumerate}
  \item \textit{Extensibility:}
    \begin{enumerate}[label*=\arabic*.]%[leftmargin=1em]
      \item clear modular architecture.
    \end{enumerate}
\end{enumerate}

The robustness of analysis is the main criterion of \porthos[2] as a verification of a bug-hunting tool.
This implies modelling the system with only over-approximating abstractions, which, however makes it more vulnerable to the combinatorial growth of the state space size, preserve the completeness of analysis.
As a robust and transparent tool, the \porthos[2] must adhere to the strategy of aborting its work on any unexpected state (for instance, if a parser failed to parse the string and recovery algorithm is not described).
%The modular architecture of \porthos[2] allow each its module to be tested independently.
%Usage of immutable data structures for all DTOs helps 
%immut: preservation of invariants
%The transparency, efficiency and extensibility are the necessary criteria of any good tool.

As its predecessor, \porthos[2] uses the open-source SMT solver \tool{Z3} by Microsoft Research~\cite{de2008z3}. However, unlikely its predecessor, the \porthos[2] has an additional abstraction level \zformula{} (see Section~\ref{ch:impl:model:zformula}) that allow to use any other SMT solver. %TODO: *should* allow?

@@

Each memory unit has an \texttt{XType} associated with it.
The X-type is a symbolic representation of the C primitive type%
\footnote{Here we should note the \porthos[2] can eventually evolve to be able to analyse programs written in an OOP language (for instance, in C++). In this case, the \texttt{XType} will have more complex structure than a simple enumeration, which it has when we need to emulate only primitive types of C language. See more detailed discussion on input language type system in Section~\ref{ch:impl:proc:x-pre-compiler:type}.} %
that is easily convertable to an SMT type (modelled as \texttt{ZType}).%TODO: add the class ZType

@@

  \1 \texttt{XComputationEvent} \\
    We distinguish two types of computation events:
      %\begin{itemize}[noitemsep]
      \2 \texttt{XUnaryComputationEvent} that encodes bit negation and no-operation, 
      \2 \texttt{XBinaryComputationEvent} that encodes numeric operations (such as addition, multiplication, etc.), bit vector operations (such as bit-and, bit-xor, etc.), relative operations (such as greater-then comparison, equality comparison, etc.), and logical operations (such as conjunction and disjunction). 
      %\end{itemize}
    \newline
    The computation event class implements both \texttt{XEvent} and \texttt{XMemoryUnit}.
    This is a model-level optimisation, which is possible because a computation event performs computation over local-only memory and does not change value of any memory unit.
    Thus, the \textit{computation} abstraction (as the CPU time spent for the computation itself) can be safely removed from the model, and the computation event can be seen as a zero-time operation that produces the \textit{value}.
    For analysing large expressions this optimisation is sensible because it considers the whole expression as a single computation event, encoded therefore as a single SMT-variable.
    \newline
    Note, for the purpose of simplifying the X abstraction level, computation events may be modelled as a bodiless function invocation (for instance, `\lstinline{x + y}' is the invocation of the function `\lstinline{+}' to argument `\lstinline{x}' and `\lstinline{y}').

  \1 \texttt{XControlFlowEvent} \\
    The control-flow event indicates a non-linear jump in the code. We distinguish two kinds of control-flow events: 
    %\begin{itemize}[noitemsep]
      \2 \texttt{XJumpEvent} that performs no computation and no data operation, it can be safely removed from the model as an optimisation,
      \2 \texttt{XMethodCallEvent} that models the method call with the \texttt{FASTCALL} calling convention (passing arguments in registers).
      %TODO: check the diagrams if the XMCE really inherits XLMU
      The method call also implements the \texttt{XLocalMemoryUnit} since it represents the computed value returned by the method call.
      If the method has been resolved, the actual arguments are bind to the formal parameters (treated as temporary local memory units), the called \texttt{XMethodCallEvent} is pushed onto the call stack, and the execution jumps to the method body.
      Each \texttt{return} statement creates the assignment of the method call event on the top of call stack.
      \\
      If the method has not been resolved, \textbf{TODO: think about it. Will the formula be SAT? Or do we need to abort the analysis?}
    %\end{itemize}

@@

\begin{figure}[h]
\centering
%\begin{multicols*}{2}
\begin{minipage}[t]{.55\textwidth}
\begin{lstlisting}[language=java]
interface Element {
  <T> T accept(Visitor<T> visitor);
  ...
}

class AnElement implements Element {
  @Override
  <T> T accept(Visitor<T> visitor) {
    return visitor.visit(this);
  }
  ...
}
\end{lstlisting}
%[* \vfill\null \columnbreak *]
\end{minipage}% <---------------- Note the use of "%"
\begin{minipage}[t]{.45\textwidth}
\begin{lstlisting}[language=java]
class Visitor<T> {
  T visit(AnElement e) {
    // visiting logic
    ...
    // continue recursively
    e.getChild().accept(this);
  }
  T visit(AnOtherElement e) {
    ...
  }
  ...
}
\end{lstlisting}
%\end{multicols*}
\end{minipage}
\caption{Illustration of the visitor pattern}
\label{fig:visitor}
\end{figure}

@@

Both input-language and input-model parsers are implemented via ANTLR parser generator.

Details on the input program language grammar are discussed in Section~\ref{ch:impl:input}.
Currently, \porthos[2] does not consider preprocessor instructions (it ignores them).
For implementing the inter-procedural mode of \porthos{}, it is crucial to support inclusion of header files (the \texttt{\#include} preprocessor directive).
Next step would be implementing the support of macros and conditional compilation directives, that are used often in C code.
As preprocessor statements may appear in arbitrary place of a program, the preprocessor must be a stateful processing unit that reads the token stream and dynamically instrument the program by interpreting directives and expanding macros.
%The ANTLR grammar for the C preprocessor may be found 
%more macros (stateful) : https://gcc.gnu.org/onlinedocs/cpp/Macros.html

The input memory model language \cat is discussed in Section~\ref{ch:wmm:cat}.
The ANTLR grammar for the \cat language was extracted from the parser used by \tool{herd} tool%
\footnote{\tool{herd} project repository: \url{https://github.com/herd/herdtools7}} %
written in OCaml.
%TODO: upload the grammar to github
%The input memory model 
\textbf{TODO: add reference to the grammar file in github repository OR include a short version as an appendix}

@@

todo: file tree of w-model to appendix

@@

The precompiler traverses the \ytree{} and collects information necessary for compiling it into an \xgraph{}.

%TODO: not supported yet

The label resolution is necessary for establishing links to labelled statements. %a process of linkage the text labels referenced in code to the places of their declaration.
In C, labelled statements are declared via the colon-syntax `\texttt{<label> : <statement>}', and the labels are referenced by the jump-statement `\texttt{goto <label>}'.
The label resolution algorithm traverses the \ytree{} and collects all declared labels into a map that points a label to the labeled statement. %TODO: code: remove temp label generation, it's stupid. put null.
This information is used during compilation to set up unconditional jumps.

\paragraph{Type analysis}
\label{ch:impl:proc:x-pre-compiler:type}

The C language has a static (resolved at compile-time) manifest (all types are declared explicitly) type system.
Comparing to languages that use type inference, the type analysis of a C program constitutes a simple propagating the type information (obtained from variables declarations) to all expressions. % according the rules of propagation. For example, 
Being carried at Y and X representation levels, the type is converted to a Z-type at the stage of the Z-formula encoding (see Section~\ref{ch:impl:proc:z-encoder}).
Currently, \porthos[2] operates only the primitive C types (such as \texttt{int}, \texttt{char}, \texttt{float}, etc.), which is modelled as an enumeration \texttt{XType}. %TODO: model it.
Array dimension is stored as an integer number at the Y-level (at the X- and Z- levels, \textbf{TODO: think about it}). %the array operations are converted to ).
In addition, \porthos[2] supports ... int\_t ....

   %TODO: support this.

%TODO: describe the XType
%TODO: definitely get rid of bit32

%TODO: <not implemented>
The type analysis algorithm should consider the type aliases supported by C language (defined by the \texttt{typedef} instruction).
For resolving the type aliases, the precompiler should make an extra traverse of the \ytree{} before the precompilation stage  and build up a symbol map.

%TODO: <not implemented>
The type analysis includes the function resolution: for correct resolving the semantics of functions, the compiler needs a mapping of each function invocation to its signature.
However due to lack of polymorphism and function overloading mechanism inherent to OOP languages the function resolution algorithm may set up the mapping only for the function name and thus may neglect analysis of types of the arguments, we build the mapping for the full function signature so that the type analysis preprocessing unit may be used for analysing the C++ code if needed.

@@

As it was discussed in Section~\ref{ch:impl:model:ytree}, the \ytree{} encodes all named objects as \texttt{YVariableRef}.

@@

@@

mapping
